---
title: "Task 1: Principal Component Analysis"
author: "Claire Meuter"
date: "2023-03-03"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) #all hail the tidyverse
library(here) #easy reading
library(ggfortify) # biplot creation
```

# Introduction
For task 1, I will be exploring pattern within a data set on environmental and climatic variables (at a country-level). This data was compiled and provided by @zander_venter on Kaggle, described as: “This data is all acquired through Google Earth Engine (https://earthengine.google.com/) where publicly available remote sensing datasets have been uploaded...Most of the data is derived by calculating the mean for each country at a reduction scale of about 10km.”

I will be exploring this data using a Principal components analysis (PCA). What is a PCA? A PCA reprojects original axes in a data set to new principal components that capture the greatest amount of variance in a multivariate data set. PCA results can be visualized as biplots and screeplots. 

# Data wrangling 
To prepare my data for a PCA, will drop categorical and character data, and exclude any rows with NAs. I should also rescale so all numeric variables have a mean 0 and sd 1.

First, I read in my data and drop rows with NAs, and drop categorical and character data. Then, I'll rescale all of my numeric variables to have a mean of 0 and a standard dev of 1.
```{r}
#reading in my data set
world_data <- read_csv(here("data", "world_env_vars.csv")) %>%
  drop_na() %>% #dropping any rows containing NAs
  dplyr::select(where(is.numeric)) %>% #keeping only numeric columns
  dplyr::select(-aspect, -slope) %>% # there's a lot of variables here. I'm going to drop some that I just don't think are very informative 
  dplyr::select(-starts_with(c('rain','temp')))# I'm dropping all the rain ones 

world_pca <- world_data %>%  
  scale() %>% #putting all my variable #s on a similar scale 
  prcomp() # creating PCA


# A little data exploration first: 
#unique(world_vars$Country) #this wont work once I've selected for numeric variables only 
```
# See the loadings (weighting for each principal component)

```{r}
world_pca$rotation
```

# Visualize PCAs 
Now that my PCAs, I have a few options on how to display this data. I can create biplot loadings by hand, or I can just use autoplot(). Autoplot is faster, and I'm a busy grad student, so that'll be my choice. 

Here, I create a biplot to look at the PCA results in a two-dimension format
```{r}
autoplot(world_pca, 
         data = world_data,
         loadings = TRUE,
         loadings.label = TRUE,
         loading.colour = "black",
         loadings.label.colour = "black") +
  theme_minimal()
```

