---
title: 'Task 2: Agglomerative hierarchical clustering'
author: "Claire Meuter"
date: "2023-03-05"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) #all hail the tidyverse
library(dendextend) #tanglegram
library(ggdendro)
library(here) #easy reading

```

## Introduction
In this analysis, I will be using agglomerative hierarchical clustering to explore a data set on stream chemistry in the Santa Barbara Coastal drainage area. This analysis treats each object (in this case, stream sites) as it's own cluster. Next, pairs of clusters are successively merged together based on their average chemical similarity. The output from this analysis is a tree-based representation called a dendrogram. The chemical variables analysed here are dissolved nitrogen, soluble reactive phosphorous, particulate organic carbon, total suspended solids, and specific conductivity. 

**Reference:** Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174. 

## Data wrangling
Here, I'll read in the raw data and prepare it for a hierarchical clustering. I'm going to drop any variables with a high (>50%) occurrence of NAs. I convert any -999 values to NA. I will rename site locations. I'll evaluate for and delete any variables that have more than a 50% NAs. I will find the average stream chemistry values for each site, and the scale the data to easily calculate euclidean distances. 

```{r}
#read in the data
stream <-
  read_csv(here("data","sbc_lter_registered_stream_chemistry.csv")) %>% 
  na_if(-999) %>%  #assigning na to -999 values 
  mutate(site_name = case_when(site_code == 'AB00' ~ 'Arroyo Burro',
                               site_code == 'AT07' ~ 'Atascadero Creek',
                               site_code == 'BC02' ~ 'Bell Canyon Creek',
                               site_code == 'DV01' ~ 'Devereaux Creek',
                               site_code == 'GV01' ~ 'Gaviota',
                               site_code == 'HO00' ~ 'Arroyo Hondo',
                               site_code == 'MC00' ~ 'Mission Creek (Montecito St)',
                               site_code == 'MC06' ~ 'Mission Creek (Rocky Nook)',
                               site_code == 'ON02' ~ 'San Onofre Creek',
                               site_code == 'RG01' ~ 'Refugio Creek',
                               site_code == 'RS02' ~ 'Rattlesnake Creek',
                               site_code == 'SP02' ~ 'San Pedro Creek',
                               site_code == 'TO02' ~ 'Tecolote Creek'), 
         .after = site_code) #renaming the sites 



### explore the number of NAs by variable
 #colSums(is.na(stream))
 #  site_code         site_name   timestamp_local            nh4_uM 
    #            0                 0                 0               203 
  #         no3_uM            po4_uM            tdn_uM            tdp_uM 
   #           149               167              5034              5801 
   #        tpc_uM            tpn_uM            tpp_uM    tss_mgperLiter 
   #         16521             16521             17046             14376 
#spec_cond_uSpercm 
  #            924 

#There are 19,390 obs, so I want to drop any variables that have more than 9,695 NAs. 
# - tpc-uM, tpn_uM, tpp_uM, tss_mgperLiter all have more than 50% NAs
 
### I'll drop them from the analysis
### Next, I find the average values by stream site.


stream_avgs <- stream %>% 
  select(!c(tpc_uM, tpn_uM, tpp_uM, tss_mgperLiter)) %>% #dropping unwanted variables
  drop_na() %>% 
  group_by(site_name) %>% 
  summarize(nh4_mean = mean(nh4_uM, na.rm = TRUE),
            no3_mean = mean(no3_uM, na.rm = TRUE),
            po4_mean = mean(po4_uM, na.rm = TRUE),
            tdn_mean = mean(tdn_uM, na.rm = TRUE),
            tdp_mean = mean(tdp_uM, na.rm = TRUE),
            spec_cond_mean = mean(spec_cond_uSpercm, na.rm = TRUE))
### Next I scale the numeric data
stream_chem_scale <- stream_avgs %>% 
  select(where(is.numeric)) %>% 
  scale()
### reassign rownames based on site name
rownames(stream_chem_scale) <- stream_avgs$site_name
```

## Hierarchical clustering
Hierarachical clustering can be top-down, or bottom-up. As stated above, am agglomerate method means that we take a bottom-up approach. In other words, our stream sites are each treated as their own cluster, and we now need to group them together based on similarity. 
To do this, I need to calculate the Euclidean distance in multivariate space between the different observations. This matrix of distances will allow me perform hierarchical clustering. These results can be displayed by a dendrogram.  

```{r}
### calculate the euc dists
euc_dist <- dist(stream_chem_scale, method = 'euclidean')
### use stats::hclust() for hierarchical clustering
stream_hc_complete <- hclust(euc_dist, method = 'complete')
### quick base plot check:
 plot(stream_hc_complete, cex = 0.6, hang = -1)
```

**Figure A.** A dendrogram showing Santa Barbara streams chemistry similarity by hierarchical clustering. The y-axis indicates how similar the stream chemistry is by water sample sites. 

## Summary 
Our cluster dendrogram can give us insights into the similarities between streams in Santa Barbara based on water samples. 

 - San Pedro Creek and Mission Creek (Rocky Nook) share the most similarities, on average
 - Devereaux Creek is the most dissimilar of the streams 
