---
title: 'Task 2: Agglomerative hierarchical clustering'
author: "Claire Meuter"
date: "2023-03-05"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse) #all hail the tidyverse
library(dendextend) #tanglegram
library(ggdendro)
library(here) #easy reading

```

# Introduction
In this analysis, I will be using agglomerative hierarchical clustering to explore a data set on stream chemistry in the Santa Barbara Coastal drainage area. This analysis treats each object (in this case, stream sites) as it's own cluster. Next, pairs of clusters are successively merged together based on their average chemical similarity. The output from this analysis is a tree-based representation called a dendrogram. The chemical variables analysized here are dissolved nitrogen, soluble reactive phosphorous, particulate organic carbon, total suspended solids, and specific conductivity. 


**Source:** Citation: Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174. 

# Data wrangling
Here, I'll read in the raw data and prepare it for a hierarchical clustering. I'm going to drop any variables with a high (>50%) occurance of NAs. I convert any -999 values to NA. I'll evaluate for and delete any variables that have more than a 50% NAs. I will find the average stream chemistry values for each site, and the scale the data to easily calculate euclidean distances. 

```{r}
stream_chem <-
  read_csv(here("data","sbc_lter_registered_stream_chemistry.csv")) %>% 
  na_if(-999) %>% 
  mutate(site_name = case_when(site_code == 'AB00' ~ 'Arroyo Burro',
                               site_code == 'AT07' ~ 'Atascadero Creek',
                               site_code == 'BC02' ~ 'Bell Canyon Creek',
                               site_code == 'DV01' ~ 'Devereaux Creek',
                               site_code == 'GV01' ~ 'Gaviota',
                               site_code == 'HO00' ~ 'Arroyo Hondo',
                               site_code == 'MC00' ~ 'Mission Creek (Montecito St)',
                               site_code == 'MC06' ~ 'Mission Creek (Rocky Nook)',
                               site_code == 'ON02' ~ 'San Onofre Creek',
                               site_code == 'RG01' ~ 'Refugio Creek',
                               site_code == 'RS02' ~ 'Rattlesnake Creek',
                               site_code == 'SP02' ~ 'San Pedro Creek',
                               site_code == 'TO02' ~ 'Tecolote Creek'), 
         .after = site_code) #renaming the sites for ease later



### explore the number of NAs by variable
 colSums(is.na(stream_chem))
 #  site_code         site_name   timestamp_local            nh4_uM 
    #            0                 0                 0               203 
  #         no3_uM            po4_uM            tdn_uM            tdp_uM 
   #           149               167              5034              5801 
   #        tpc_uM            tpn_uM            tpp_uM    tss_mgperLiter 
   #         16521             16521             17046             14376 
#spec_cond_uSpercm 
  #            924 

#There are 19,390 obs, so I want to drop any variables that have more than 9,695 NAs. 
# - tpc-uM, tpn_uM, tpp_uM, tss_mgperLiter all have more than 50% NAs
 
### I'll drop them from the analysis
### Next, I find the average values by stream site.


stream_chem_avgs <- stream_chem %>% 
  select(!c(tpc_uM, tpn_uM, tpp_uM, tss_mgperLiter)) %>% #dropping unwanted variables
  drop_na() %>% 
  group_by(site_name) %>% 
  summarize(nh4_mean = mean(nh4_uM, na.rm = TRUE),
            no3_mean = mean(no3_uM, na.rm = TRUE),
            po4_mean = mean(po4_uM, na.rm = TRUE),
            tdn_mean = mean(tdn_uM, na.rm = TRUE),
            tdp_mean = mean(tdp_uM, na.rm = TRUE),
            spec_cond_mean = mean(spec_cond_uSpercm, na.rm = TRUE))
### scale the numeric data
stream_chem_scale <- stream_chem_avgs %>% 
  select(where(is.numeric)) %>% 
  scale()
### reassign rownames based on site name
rownames(stream_chem_scale) <- stream_chem_avgs$site_name
```

# Hierarchical clustering
For a bottom-up hierarchical cluster, we first need to calculate the Euclidean distance in multivariate space between the different observations. Then, we use this matrix of distances (dissimilarity matrix) to perform a complete hierarchical clustering. The results are visualized by the dendrogram in Figure 1 below.

```{r}
### calculate euc dists
euc_dist <- dist(stream_chem_scale, method = 'euclidean')
### use stats::hclust() for hierarchical clustering
stream_hc_complete <- hclust(euc_dist, method = 'complete')
### quick base plot check:
 plot(stream_hc_complete, cex = 0.6, hang = -1)
```


```{r}
### Dendrogram plot
ggdendrogram(stream_hc_complete, 
             rotate = TRUE,
             labels = TRUE) +
  theme_classic() +
  labs(x = "Sample site",
       y = "Height") +
  scale_y_continuous(breaks = seq(0, 7, by = 1),
                     expand = c(0, 0), limits = c(0, 7)) +
  theme(
    ## x axis
    axis.line.x = element_line(linewidth = 0.8),
    axis.ticks.x = element_line(linewidth = 0.8),
    axis.ticks.length.x = unit(.25, 'cm'),
    axis.text.x = element_text(face = 'bold', size = 9.5),
    axis.title.x = element_text(face = 'bold', size = 11, vjust = -1),
    ## y axis
    axis.line.y = element_blank(),
    axis.title.y = element_text(face = 'bold', size = 11, vjust = 3.5),
    axis.text.y = element_text(face = 'bold', size = 9))
```

**Figure 1.** Santa Barbara Coastal LTER stream chemistry hierarchical clustering by complete linkage. The x-axis of this dendrogram indicates how closely related the stream chemistry is by water sample sites, on average.  

